draft 1

### 2016-06-13 finalwriteup/draft.rmd
finished rough draft for paul

opts_chunk$set(fig.width=10, fig.depth = 7, dpi = 300, warning = FALSE, message = FALSE, echo = FALSE)
```

@@

B2B- Business to Business

@@

* vague interesting few sentences - context

@@

# Chapter 1 Problem Description

@@

## 1.1 Research Motivations - 4.5 pages

description of problem:

@@

Why its not easily solvable:

@@

current practise:

@@

theory about problem with current practise:

@@

previous research:

@@

### 1.1.1 Overview of Past Reference Class Forecasting Methods and relevant Industry Uptake

@@

***need to really do more research on this!!***

#### 1.1.1.1 Existing Models

@@

#### 1.1.1.2 Industry Uptake of Existing Models and Current Practises

@@

### 1.1.2 Case Study

@@

### 1.1.3 Case Study Limitations

@@

### 1.1.4 Research Motivation Summary

@@

## 2.0 Thesis aims

@@

## 1.2 Thesis contributions - Gaps in literature

@@

* the application of clustering methods to business to business decisions.

## 1.3 Thesis structure

lit review
Background info
present the case study
Present the models you've developed
summary of findings, implications for use within industry, limitations

# Chapter 2 Literature Review - 25 pages for pHd

@@

## 2.1 Project Cost Estimation Methods

@@

### 2.1.1 Cost Estimation in the Construction Industry

@@

**summary**

@@

### 2.1.1 Effort Estimation in the IT Industry

@@

## 2.2 Statistical and machine learning methods

### 2.2.1 Introduction

@@

### 2.2.2 Linear Regression

@@

### 2.2.3 Logistic Regression

@@

### 2.2.4 Naive Bayes

@@

### 2.2.5 Decision Trees/ Ensemble Trees

@@

### 2.2.5 Bayesian Networks

@@

### 2.2.6 SVM's and Neural Networks

@@

### 2.2.7 Summary

@@

## 2.3 Statistical and machine learning methods applied to business problems

@@

### 2.3.2 Summary of Advanced Statistical and Machine Learning Methods Applied to Business Problems

@@

### 2.3.3 Employee Churn Case Study

@@

### 2.3.4 Gap/conclusion

@@

## 2.4 Conclusion

@@

# Chapter 3 Methods - 2 to 8 pages

@@

**justify the process by which the research questions were answered**
**argue and justify each decision that was taken to arrive at how to organise reserach**

## 3.1 Obtaining data

@@

## 3.2 Cleaning and variable engineering

@@

* invoicing data
* timesheet data
* project summary data

@@

### 3.2.1 Variable Engineering

@@

## 3.3 Variable selection

### 3.3.1 Outlier deletion

@@

### 3.3.2 Variable selection Methods

@@

#### 3.3.2.1 ANOVA

@@

#### 3.3.2.2 Random Forest

@@

#### 3.3.2.3 Cforests

@@

#### 3.3.3 Variable Selection Summary

@@

## 3.4 Model selection

@@

## 3.5 Model comparison, Model averaging

@@

## 3.6 Bottom Line Translation

@@

## 3.7 Method Conclusion

In summary, to determine whether a 
* obtained data: 4169 jobs from a single case study company over 12 years. Data was de-dentified by an employee of the company
* the data was all from the company CRM. There were three chief sources:
    * invoicing data
    * timesheet data
    * project summary data
* the data was cleaned from mistakes which were generally discovered as outliers 
* many new variables were engineered from the data, often by summarising time series data into a value that is relevant to a final data set of a single job per row
* variable selection: dozens of variables available which were condensed to the most important 10-12. variable importance methods were ANOVA, random forest, and cforests. the overlapping important variables were chosen first as well as some others
* 5 models were tested: baseiline: log reg, naiive bayes.
* sophisticated ML: boosted reg, random frest, bayesian networks
* best were comapred to a model averaged version
* Analysis of impact on bottom line for business if model is integrated into decision making
* link: results

# Chapter 4: Results

@@

### cForest

@@

### Variable selection summary

@@

mice.totAUC <- readRDS("report_data/mice_df_auc.rds")

@@

cfor.varimp = readRDS("report_data/cfor_blend_varimp.rds")

@@

log.varimp = readRDS("report_data/logblend.rds")

@@

rf.varimp = readRDS("report_data/rfblend.rds")
rf.varimp = importance(rf.varimp, type = 1) %>% as.data.frame
rf.varimp = add_rownames(rf.varimp)
rf.varimp = arrange(rf.varimp, desc(MeanDecreaseAccuracy))
# order levels
rf.varimp$rowname <- factor(rf.varimp$rowname,
                               levels= rf.varimp$rowname)

@@

boost.varimp = readRDS("report_data/boostblend.rds")
boost.varimp = add_rownames(boost.varimp)
# order levels
boost.varimp$rowname <- factor(boost.varimp$rowname,
                               levels= boost.varimp$rowname)

@@

auc.list <- readRDS("report_data/blend_auc1.rds")
means = lapply(auc.list, mean) %>% unlist
greater = means[means> means["orig.log"]] %>% names
auc.melt = melt(auc.list)
auc.melt$greater = ifelse(auc.melt$L1 %in% greater, TRUE, FALSE)

@@

```{r}
#load results from 150 iterations
auc.150 = readRDS("report_data/blend_auc150.rds")

@@

The results of the 150 iterations are shown in the plot below. Simple model averaging results in a p-value of `t.test(auc.150[['orig.log']], auc.150[['average']])$p.value %>% format(digits = 2) %>% as.numeric` while the simple logisitc regression blended model had a two-sample t.test p-value of `t.test(auc.150[['orig.log']], auc.150[['simp.log']])$p.value %>% format(digits = 2) %>% as.numeric`.

@@

```{r profit_curve, fig.depth = 10}
finance.plot = readRDS("report_data/plot_finance.rds" )

@@

```{r av_profit_curve}
ggplot(finance.plot %>% filter(method == "average"), 
       aes(x= threshold, y= profit.ratio, colour = method)) + 
        geom_point(alpha = 0.1, shape = 16) + 
        stat_smooth() +
        scale_y_continuous(breaks = seq(0,110, by = 10)) + 
        scale_x_continuous(breaks = seq(0,1, by = .1)) + 
        geom_hline(yintercept = 100, colour = "darkgray", linetype = 2) + 
        geom_hline(yintercept = 0, colour = "darkgray", linetype = 1) + 
        theme(legend.position = "none") +
        labs(x = "Probability Threshold", 
             y = "Ratio of Model Profit to Historical Profit (%)",
             title = "Profit Curve for Blended Average Model")

@@

# Results - over 5 pages, up to 20 pages

Guidelines:
* Outlines what you found out in relation to your research questions or hypotheses, presented in figures and in written text.
* Results contain the facts of your research. Often you will include a brief comment on the significance of key results, with the expectation that more generalised comments about results will be made in the Discussion section.

Which results to discuss
* restate research questions and hypotheses. can we predict profitability? ofocurse you can make a prediction, 
real question might be is it worthwhile for the business to engage in this activity? we will look at this as a bottom line analysis. acknowledge that it may be worthwhile for other reasons such as confidence and transparency in historic projects.

* bottom line analysis - run 1000 times to get confidence interval around the line.
    * should the business adopt the model?

* to address the problem of trust and create insight to the past data - nearest neighbour algorithm. This also mimics the case based reasoning approach

@@

The Discussion section:

comments on your results;
explains what your results mean;
interprets your results in a wider context; indicates which results were expected or unexpected;
provides explanations for unexpected results.
The Discussion should also relate your specific results to previous research or theory. You should point out what the limitations were of your study, and note any questions that remain unanswered. The Discussion CAN also include Conclusions/Future Research.

@@

simp.l <- readRDS("report_data/finance_average_model.rds")

@@

Because of the positive AUC results, it was expected the models would translate into financial benefits for the company. The results look reasonable, profit gains may be high enough to trigger further cost-benefit analyses for introducing the model into business decisons. It was surprising that, given the similar AUC results, simple model averaging signficantly (?) outperformed the blended logistic regression model's profit curve. This may be because ... ROC curves.

*** you are here ****

Previous research indicated problems with user uptake once models were developed and a lack of trust in the model output [@Akintoye2000].

* problems with past research and does this solve any of those?
    * poor user uptake - future work?
    * lack of trust - nearest neighbour must be presented with results - case based reasoning
    * results from boosted tree shows partial dependence plots - insight.
    * never been applied to a consultant within construction industry. has been applied lots to software
    * still has the potential to help with the way the industry is structured - hinges on user uptake.

You should point out what the limitations were of your study, and note any questions that remain unanswered. 
* not sure how easy number of users is to predict - could do a study
* could do a study on trust levels

@@

Future work
* user testing on trust levels of prediction with nearest neighbour

Conclusions
could be a viable business move worth integrating

Suggestions

Could collect data for future projects: expected team size, expected time frame.

* trialled ensemble tree methods on cost estimation (in this case project proftability)
* attempted to predict project profitability using a single consulting company's CRM data in the construction industry
* assessed tangible profit outcomes of machine learning models

>>>>

opts_chunk$set(fig.width=8, fig.depth = 8, dpi = 300, warning = FALSE, message = FALSE, echo = FALSE)
```

@@

B2B - Business to Business
CRM - Customer Relationship Management. Refers to a common type of software used by businesses to record client and project details as well as employee timesheet records. Generally accessable to each employee in a company.

@@

@@

# Problem Description

@@

## Research Motivation - 4.5 pages

**Problem Description:**

@@

**Why its Not Easily Solvable:**

@@

**Current Practise:**

@@

@@

@@

### Overview of Past Reference Class Forecasting Methods and relevant Industry Uptake

@@

#### Existing Models

@@

#### Industry Uptake of Existing Models and Current Practises

@@

### Case Study

@@

### Case Study Limitations

@@

### Research Motivation Summary

@@

## Thesis aims

@@

## Thesis Contributions - Gaps in literature

@@

# Literature Review - 25 pages for pHd

@@

## Project Cost Estimation Methods

@@

### Cost Estimation in the Construction Industry

@@

@@

### Effort Estimation in the IT Industry

@@

## Statistical and machine learning methods

### Introduction

@@

### Linear Regression

@@

### Logistic Regression

@@

### Naive Bayes

@@

### Decision Trees/ Ensemble Trees

@@

### Bayesian Networks

@@

### SVM's and Neural Networks

@@

### Summary

@@

## Statistical and Machine Learning Methods Applied to Business Problems

@@

### Summary of Advanced Methods

@@

### Employee Churn Case Study

@@

### Gap

@@

## Conclusion

@@

# Chapter 3 Method - 2 to 8 pages

@@

## Obtaining data

@@

## Cleaning and variable engineering

@@

* Invoicing data
* Timesheet data
* Project summary data

@@

### Variable Engineering

@@

### Outlier Deletion

@@

### Variable Selection Methods

@@

@@

#### Random Forest

@@

#### Cforests

@@

#### Variable Selection Summary

@@

## Model selection

@@

## Model Comparison, Model Averaging

@@

## Bottom Line Translation

@@

```{r}
## Method Conclusion

# In summary, to determine whether a 
# * obtained data: 4169 jobs from a single case study company over 12 years. Data was de-dentified by an employee of the company
# * the data was all from the company CRM. There were three chief sources:
#     * invoicing data
#     * timesheet data
#     * project summary data
# * the data was cleaned from mistakes which were generally discovered as outliers 
# * many new variables were engineered from the data, often by summarising time series data into a value that is relevant to a final data set of a single job per row
# * variable selection: dozens of variables available which were condensed to the most important 10-12. variable importance methods were ANOVA, random forest, and cforests. the overlapping important variables were chosen first as well as some others
# * 5 models were tested: baseiline: log reg, naiive bayes.
# * sophisticated ML: boosted reg, random frest, bayesian networks
# * best were comapred to a model averaged version
# * Analysis of impact on bottom line for business if model is integrated into decision making
# * link: results
```

# Results

@@

### Cforest

@@

### Variable Selection Summary

@@

mice.totAUC <- readRDS("mice_df_auc.rds")

#significant difference??
index = c(1,3,4)
pvals = data.frame("method" = c('log','rf','naiive'), 'p_value' = NA)
for(i in seq_along(index)){
        pvals$p_value[i] = t.test(mice.totAUC[[2]], mice.totAUC[[index[i]]])$p.value %>%
                format(digits = 2) %>% as.numeric
}

signif = pvals %>% filter(p_value < 0.05) %>% .$method
mice.totAUC = melt(mice.totAUC)
mice.totAUC$sigdiff = ifelse(mice.totAUC$L1 %in% signif, TRUE, FALSE)

ggplot(mice.totAUC, aes(x= factor(L1), y= value)) + 
        geom_violin(aes(fill = sigdiff, colour = sigdiff)) + 
        geom_jitter(alpha = 0.2, shape = 16, colour= 'gray70') +
        stat_summary(fun.y="mean", geom="point", colour = "gray60", shape =3) +
        scale_fill_manual(values = c('sienna1','navajowhite'),
                          name = "Significantly different \nfrom boosted trees") +
        scale_colour_manual(values = c('sienna1','navajowhite'),
                          name = "Significantly different \nfrom boosted trees") +
        labs(x = "Method", y = "Area Under Curve (AUC)",
             title = "AUC Distribution from 100 models of each Method - Imputed Data") +
        scale_x_discrete(labels = c("boost" = "Boosted Trees", 
                         "log" = "Logistic Regression",
                         "naiive" = "Naiive Bayes", "rf" = "Random Forest"))
```

@@

cfor.varimp = readRDS("cfor_blend_varimp.rds")

@@

log.varimp = readRDS("logblend.rds")

@@

rf.varimp = readRDS("rfblend.rds")
rf.varimp = importance(rf.varimp, type = 1) %>% as.data.frame
rf.varimp = add_rownames(rf.varimp)
rf.varimp = arrange(rf.varimp, desc(MeanDecreaseAccuracy))
# order levels
rf.varimp$rowname <- factor(rf.varimp$rowname,
                               levels= rf.varimp$rowname)

@@

boost.varimp = readRDS("boostblend.rds")
boost.varimp = add_rownames(boost.varimp)
# order levels
boost.varimp$rowname <- factor(boost.varimp$rowname,
                               levels= boost.varimp$rowname)

@@

```{r, fig.height=10}

auc.list <- readRDS("blend_auc1.rds")
means = lapply(auc.list, mean) %>% unlist
greater = means[means> means["orig.log"]] %>% names
auc.melt = melt(auc.list)
auc.melt$greater = ifelse(auc.melt$L1 %in% greater, TRUE, FALSE)

@@

```{r}
#load results from 150 iterations
auc.150 = readRDS("blend_auc150.rds")

@@

The results of the 150 iterations are shown in the plot below. Simple model averaging results in a p-value of `r t.test(auc.150[['orig.log']], auc.150[['average']])$p.value %>% format(digits = 2) %>% as.numeric` while the simple logisitc regression blended model had a two-sample t.test p-value of `r t.test(auc.150[['orig.log']], auc.150[['simp.log']])$p.value %>% format(digits = 2) %>% as.numeric`.

```{r, fig.width = 8, fig.height=10}

@@

```{r profit_curve, fig.depth = 10}
finance.plot = readRDS("plot_finance.rds" )

@@

```{r av_profit_curve, fig.height=10}
ggplot(finance.plot %>% filter(method == "average"), 
       aes(x= threshold, y= profit.ratio, colour = method)) + 
        geom_point(alpha = 0.1, shape = 16) + 
        stat_smooth() +
        scale_y_continuous(breaks = seq(0,110, by = 10)) + 
        scale_x_continuous(breaks = seq(0,1, by = .1)) + 
        geom_hline(yintercept = 100, colour = "darkgray", linetype = 2) + 
        geom_hline(yintercept = 0, colour = "darkgray", linetype = 1) + 
        theme(legend.position = "none") +
        labs(x = "Probability Threshold", 
             y = "Ratio of Model Profit to Historical Profit (%)",
             title = "Profit Curve for Blended Average Model")

@@

```{r}
# Results - over 5 pages, up to 20 pages
# 
# Guidelines:
# * Outlines what you found out in relation to your research questions or hypotheses, presented in figures and in written text.
# * Results contain the facts of your research. Often you will include a brief comment on the significance of key results, with the expectation that more generalised comments about results will be made in the Discussion section.
# 
# Which results to discuss
# * restate research questions and hypotheses. can we predict profitability? ofocurse you can make a prediction, 
# real question might be is it worthwhile for the business to engage in this activity? we will look at this as a bottom line analysis. acknowledge that it may be worthwhile for other reasons such as confidence and transparency in historic projects.
# 
# * bottom line analysis - run 1000 times to get confidence interval around the line.
#     * should the business adopt the model?
#     
# * to address the problem of trust and create insight to the past data - nearest neighbour algorithm. This also mimics the case based reasoning approach
```

@@

```{r}
# The Discussion section:
# 
# comments on your results;
# explains what your results mean;
# interprets your results in a wider context; indicates which results were expected or unexpected;
# provides explanations for unexpected results.
# The Discussion should also relate your specific results to previous research or theory. You should point out what the limitations were of your study, and note any questions that remain unanswered. The Discussion CAN also include Conclusions/Future Research.
```

@@

simp.l <- readRDS("finance_average_model.rds")

@@

Because of the positive AUC results, it was expected the models would translate into financial benefits for the company. The results look reasonable, profit gains may be high enough to trigger further cost-benefit analyses for introducing the model into business decisons. It was surprising that, given the similar AUC results, simple model averaging signficantly (?) outperformed the blended logistic regression model's profit curve. This means that model averaging predicted projects with probabilities closer to 1 (loss-making jobs) with more accuracy than blended logistic regression. This may be because ... ROC curves.

### Limitations and Future Work

One of the unknown factors in translating this research into business practises is how well users can guess the project variables. The data used to develop the model was obtained from the company's records of how projects played out and there are no records of initially expected values of variables. For example, if a project lasted 2 years and it was initially expected it would only take 1.5 years, the only data the company retains is that the project lasted two years. This begs the question, how well could a user predict the variables before beginning a project? Uncertainty has been managed to a degree by binning variables such as 'total amount invoiced' and 'timespan' to 5 or 6 broad categories (i.e. "1 day - 3 weeks", "3 weeks - 2.5 months", etc). However, two remaining categories: team size and percent of project completed by professional level employees were left as continuous variables. It would be interesting to test how well business managers could guess these four variables at the start of a project.

The process of guessing varibles at the beginning of a project and then waiting for the project to complete is a long checking process. However, it would be worthwhile integrating this data into the system's pre-existing data collecting processes. When a new project is entered into the database, it could be required that the project manager guesses the time-span (1 of 6 categories), total amount invoiced (1 of 5 categories), team size and percent to be completed by professionals. This data itself could feed into a revised version of the prediction algorithm.

Another limitation for these types of algorithms in general, as identified in the literature, is user uptake of the models and a lack of trust in the output [@Akintoye2000]. It was for this reason the case study only tested models that provided insight into variable relationships (ensemble trees and logistic regression) as opposed to neural networks and SVM's that operate as 'black box' predictors. It is envisioned that a future user interface for this algorithm would display variable relationships such as this partial dependency plot produced by a boosted tree model that displays the probability of losing money vs. the team size:

![](images/boosted tree/partial_dep_plots/no_users.png)

Running a nearest-neighbour algorithm concurrently with the probability algorithm could provide a cross check of the algorithm's results and engage the user's past experiences in similar projects. This could work by presenting a graphical summary of the 4 or 5 most-similar past projects to the new project being assessed by the manager. Memories of how these similar projects played out would be triggered as well as communication with other manager's that led the similar projects. This would allow the user to connect specific past experience with the probability model output. Further development and testing is required to assess how much users trust the algorithm output and how trustworthiness may be improved.

![](images/app/NN_wharf.png)

@@

The aim of this project was to predict the profitability of projects for consulting businesses using their internal CRM data. Specifically, a statistical or machine learning model was created based on historical project data. The second aim was to test whether the model would have a positive impact on the overall profitability of a case study business. These research aims were fulfilled and the prediction model was refined to a binary classification problem of whether a project would be profitable or loss-making. This suited the limited data size best which was also highly complex. After testing several machine-learning and statistical models, a blended model of logistic regression, random forests, and boosted tree models predicted project profitability best. When applied to improving overall profitability of the business, the blended model which simply averaged the raw probability results from the three aforementioned models outperformed the individual models as well as the logistic regression blended  model. Final results showed the model could improve profitability of the business by between `r (-1.96*best.sd + best.mean) %>% round(1)`% and `r (1.96*best.sd + best.mean) %>% round(1)`% (the 95% confidence interval) by rejecting projects that have a 'probability' of being a loss-making job over `r best.t` (approximately `r ((simp.l %>% filter(mean.iter>best.t) %>% nrow)/nrow(simp.l)*100) %>% round(1)`% of projects).

These results make a significant contribution to literature in this area as they provide a clear framework for how a business can realise tangible profit improvements using a set of algorithms and project data that is passively collected by businesses. Previous studies verified algorithms' prediction accuracy in cost estimation but stopped short of how the algorithms could be implemented in every day decision making and the affect on budgets. Additionally, this research tested ensemble tree methods and furthermore blended models on the problem of cost estimation. Past research had largely experimented with regression models and to some extent neural networks and case based reasoning. The advantage of ensemble tree methods is that they provide insight into the model's reasoning while having comparable predictive power to neural networks. Having said that, a limitation to the study was that there was not sufficient data to utilise the strength of ensemble tree methods and logistic regression performed simiarly. However, the final highest performing model was a blend of regression and two ensemble tree methods.

There are several aspects of this work that should be investigated further including how well managers can estimate project input variables before a project is won. Particularly, estimations of the timespan, team size, total invoiced amount, and percentage of hours completed by professionals should be tested. These initial estimates could also be included in the each project's data entry for future analysis. Finally, research into user uptake of the model should be assessed. It is likely that in order to establish a manager's trust with the algorithm, additional model output will be required such as partial dependency plots from the boosted trees. Also, a 'nearest neighbour' algorithm could be run concurrently to show the manager outcomes of past similar projects, which should align with the model's probabiltiy prediction. Overall, this project has successfully demonstrated a blended statistical and machine learning model has the potential to aid manager's in rejecting loss-making projects to improve their business's profitability.
